preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-08 Manuscript/manuscript.Rmd
Polishing discussion

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of important 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promotors are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further out, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Likewise, disruption of the boundaries that demarcate TADs and loops have been shown to lead to cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

@@

Here we present *preciseTAD*, a random forest-based approach coupled with spatial clustering (DBSCAN) and scalable partition around medoids (CLARA) that allows for more precise prediction of TAD or loop boundaries. *preciseTAD* leverages a random forest (RF) classification algorithm built on ground truth TAD(loop) boundaries obtained from a TAD(loop) calling tool and ChIP-seq defined genomic annotations as the predictor space. We first optimized our RF model by creating a machine learning framework that systematically compared the performances of different combinations of genome binning (resolution), feature engineering procedures, and re-sampling technique. In doing so, we first introduced a new strategy for establishing boundary region prediction data referred to as *shifted binning*, in which the rows of the data matrix were controlled by the data resolution (either 5kb, 10kb, 25kb, 50kb, and 100kb) used to obtain ground truth boundaries, while assuring ground truth bounderies were located central to any given genomic bin. Furthermore, we introduced a new procedure for drawing spatial associations between boundary regions and genomic annotations, referred to as *distance*-type predictors, defined as as the $log_{2}$ genomic distance in base pairs from each genomic bin the nearest ChIP-seq peak region. We benchmarked *preciseTAD* against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. For a more in depth study, we also applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from both the GM12878 and K562 cell lines.

Our machine learning framework alone yielded many interesting results. First, we showed that RF models built using *distance*-type predictors out performed models built on previously published feature engineering techniques including overlap counts and overlap percents. Moreover, we were further able to optimize our RF model by introducing random under-sampling, a technique to correct for the severe class imbalance between TAD(loop) and non-TAD(loop) boundary regions. We showed that in our optimized RF, transcription factor binding sites provided an overwhelming amount of signal in the model, indeed out-performing histone modification- and BroadHMM-specific models, contradicting a previous study that found histone marks alone were effective in predicting TAD boundaries [@huang2015predicting]. Likewise, our RF model built using *distance*-type predictors was consistently able to identify the group of the top known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143 as being highly predictive of TAD(loop) boundary regions across both cell lines. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. This suggested that random forests, when tuned and feature engineered correctly, are highly effective in accurately predicting boundary regions, contradicting a previous study that found simple multiple linear regression out performed random forests in a TAD boundary prediction setting using both human and drosophila Hi-C data [@mourad2016computational]. Plausible reasons for their findings include lack of class balancing prior to modeling and usage of rudimentary feature engineering procedures such as overlap counts and overlap percents.

Besides balanced accuracy (BA), we investigated five other performance metrics to evaluate the optimal combination of data level characteristics for RF prediction including accuracy, area under the receiver operating characteristic curve (AUC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Supplementary Tables 6-7). The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benifits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling.

The performance of *preciseTAD* was found to be empirically better than both ARROWHEAD and Peakachu ground truth boundaries. *preciseTAD* provides bp resolution predicted boundaries, thereby alleviating resolution limitations of conventional TAD-callers. We show that, unlike ARROWHEAD, *preciseTAD* does not inflate the number of predicted boundaries, instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* boundaries were found to be more enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143. Likewise *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, further highlighting their biological significance.

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined TAD caller. Ideally we would perform simulation studies to benchmark *preciseTAD* against simulated boundaries. While there are existing methods for simulating Hi-C data sets [@forcato2017comparison; @zheng2020freehi], of which boundaries can be extracted from, to date, there are no such methods for simulating ChIP-seq data. However, given *preciseTAD* is applied on chromosome-specific sections of the linear genome, and on multiple cell lines, we feel that aggregating results across chromosomes provides strong empirical evidence of the superiority of *preciseTAD*. Another limitation is that *preciseTAD* does not provide hierarchical boundaries, but instead simply provides the bp coordinate that demarcates TADs(loops). Studies have shown evidence of the hierarchical nature to the 3D architecture of the human genome [@hansen2018recent; @weinreb2016identification; @fraser2015hierarchical]. In such a hierarchy, the boundaries provided by *preciseTAD* can be interpreted as the lowest level of boundaries.

In summary, we have shown that TAD boundary prediction is a multi-faceted problem composed of many data irregularities and characteristics. Simply quering a contact matrix ignores the fundamental roles of known molecular drivers of 3D chromatin. Instead, we propose a method for turning TAD-calling into a supervised machine learning framework built using ChIP-seq annotations. Specifically, we introduce two novel concepts known as *shifted binning* and *distance*-type predictors which we use to build random forest classification models for predicting boundary regions. We then introduce *preciseTAD*, a method for leveraging random forests to predict TAD(loop) boundaries at base pair resolution. Our method is able to bridge the resolution gap between 1D ChIP-seq annotations and 3D Hi-C sequencing data for more precise and biologically meaningful boundary identification. By releasing our method as an open source R package, we hope to provide an efficient and easy-to-use tool to further explore the 3D organization of the human genome.

>>>>

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of important 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promoters are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further out, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Likewise, disruption of the boundaries that demarcate TADs and loops have been shown to lead to cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD or loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on ground truth TAD/loop boundaries obtained from a TAD/loop calling tools and ChIP-seq defined genomic annotations as the predictor space. We benchmarked *preciseTAD* against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from both the GM12878 and K562 cell lines. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and re-sampling technique addressing class imbalance. We show that distance between boundary regions and genomic annotations coupled with random undersampling technique help to form the predictor space resulting in optimal model performance. We demonstrate that binding of four transcription factors (SMC3, Rad21, CTCF, Znf143) are sufficient for accurate TAD and loop boundary predictions. The model predicts probability for each base to be a baoundary, and, coupled with spatial clustering (DBSCAN) and scalable partition around medoids clustering (CLARA), allows for the detection of boundaries at the base-level resolution.

Our random forest framework yielded several interesting observations. First, we showed that RF models built using *distance*-type predictors out performed models built on previously published feature engineering techniques including signal strength, overlap counts and overlap percents [@mourad2016computational]. [??? Add more REFs] We further demonstrated that class imbalance hinders boundary prediction, but can be effectively addressed by a simple random under-sampling technique. The unaddressed class imbalance problem in previous studies [??? Add REFs] We showed that information about only four transcription factors (SMC3, Rad21, CTCF, Znf143) is necessary and sufficient transcription for accurate boundary prediction, out-performing histone modification- and BroadHMM-built models. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Interestingly, the same transcription factors accurately predicted both TAD and loop boundaries, suggesting similarity of the mechanisms of TAD and loop formation. Although our observations contradict a previous study showing histone marks alone were effective in predicting TAD boundaries [@huang2015predicting], our findings are backed by a large body of literature [??? Add REFs]. This suggested that the random forest model, when tuned and feature engineered correctly, is highly effective in predicting boundary regions.

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Supplementary Tables 6-7 [??? Merge them into one Excel with two sheets - same message]). The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary [??? Is it because of class imbalance?]. AUROC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well [??? So, another class imbalance problem? Define class imbalance message early, why you investigated 5 other metrics]. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values [??? So why didn't you use it?]. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling. [??? The logic is messed when reading this paragraph.]

The performance of *preciseTAD* was found to be empirically better [??? As judged by what?] than both ARROWHEAD and Peakachu ground truth boundaries. Guided by both low-resolution Hi-C data and high-resolution genome annotation data, *preciseTAD* predicts base-level resolution boundaries, alleviating resolution limitations of Hi-C data. We show that, unlike ARROWHEAD, *preciseTAD* does not inflate the number of predicted boundaries [??? Where did you show this? TBD], instead, providing only the most biologically meaningful boundaries that demarcate regions of high inter-chromosomal interactions. *preciseTAD* boundaries were found to be more enriched for known architectural transcription factors including CTCF, RAD21, SMC3, and ZNF143 [??? So what?]. Likewise *preciseTAD* boundaries were found to be more conserved between GM12878 and K562 cell lines, further highlighting their biological significance. [??? So what?]

A limitation of *preciseTAD* is that it is dependent on "ground truth" boundaries provided by a user-defined TAD caller. Given the wide variery of TAD callers and their variable performance [@Zufferey:2018aa; @forcato2017comparison], it is hard to define "ground truth" TAD boundaries. Ideally we would benchmark *preciseTAD* against simulated boundaries. While methods for simulating Hi-C data sets wish boundary annotations exist [@forcato2017comparison; @zheng2020freehi], methods for simulating the associated genomic annotations (the main component guiding *preciseTAD* predictions) are lacking. Given *preciseTAD* can be trained on the whole-genome data, we feel that the total number of TAD/loop boundaries is sufficient to guide learning of the association between genomic annotations and boundaries for precise boundary predictions.

Another limitation of *preciseTAD* is that it does not provide hierarchical boundaries, but instead provides the coordinates of boundary points and regions. Studies have shown evidence of the hierarchical nature to the 3D architecture of the human genome [@hansen2018recent; @weinreb2016identification; @fraser2015hierarchical]. In such a hierarchy, the boundaries provided by *preciseTAD* can be interpreted as the lowest level of boundaries [??? This can be debated. We are training on and predicting certain level of boundaries. Suggesting discussing that false positives detected by preciseTAD may represent boundaries at finer resolution].

In summary, we demonstrate that TAD/loop boundary prediction is a multi-faceted problem requiring consideration of multiple statistical and biological properties of genomic data. Simply considering the properties of Hi-C contact matrices ignores the fundamental roles of known molecular drivers of the 3D chromatin structures. Instead, we propose a supervised machine learning framework that leverages both Hi-C contact matrix information and genomic annotations. Our method introduces three concepts - *shifted binning*, *distance*-type predictors, and *random undersampling* - which we use to build random forest classification models for predicting boundary regions. Our method is able to bridge the resolution gap between 1D ChIP-seq annotations and 3D Hi-C sequencing data for more precise and biologically meaningful boundary identification. We introduce *preciseTAD*, an open source R package for leveraging random forests to predict TAD/loop boundaries at base resolution. We hope that *preciseTAD* will serve as an efficient and easy-to-use tool to further explore the 3D organization of the genome.
